Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-09-01 ms/main.tex
assortment highlighted; connected to Fletcher2009; revised intro and discussion; revealed a mistake in Result 3 gamma presentation

% Title page
\title{Non-Vertical Cultural Transmission and the\\ Evolution of Cooperation}

@@

% Abstract
\begin{abstract}
We construct models for the cultural evolution of cooperation in a well-mixed population with vertical, horizontal, and oblique transmission. Conditions are found for fixation of cooperation, fixation of defection, and stable coexistence between cooperation and defection, and comparisons are drawn with Hamilton's rule. A spatial model is constructed and its results are compared to the well-mixed model.
We show that assortment between cooperation and horizontal transmission facilitates the evolution of cooperation and increases the population mean fitness. 
% TODO what about oblique?
\end{abstract}

@@

\citet{Eshel1982} studied a related model for the evolution of cooperative behavior.
Their model included \emph{assortative meeting}, or non-random encounters, where a fraction $m$ of individuals in the population each interact with an individual of the same phenotype, and a fraction $1-m$ interacts  with a randomly chosen individual.  
Such assortative meeting may be due, for example, to population structure or active partner choice.
In their model, cooperative behavior can evolve if\footnote{In an extended model, which allows an individual to encounter $N$ individuals before choosing a partner, the righthand side is multiplied by $E[N]$, the expected number of encounters \citep[eq.~4.6]{Eshel1982}.
} 
\citep[eq.~3.2]{Eshel1982}
\begin{equation} \label{eq:eshel1982}
c < b \cdot m \,,
\end{equation}
where $b$ and $c$ are the benefit and cost of cooperation. 
Here $m$ in \autoref{eq:eshel1982} takes the role of the relatedness $r$ in \autoref{eq:hamilton_rule}.

Here we study the evolution of a cooperative behavior that is subject to \emph{cultural transmission}, which allows an individual to acquire attitudes or behavioral traits from other individuals in its social group through imitation, learning, or other modes of communication \citep{cavalli1981cultural,richerson2008not}.
\citet{feldman1985gene} introduced the first model for the evolution of altruism by cultural transmission.
They showed that under a combination of genetic and vertical (parent-to-offspring) cultural transmission, the conditions for the initial increase of altruism entails a modification of Hamilton's rule in the cases of parent-to-offspring or sib-to-sib altruism.
For example, if the fidelity of cultural transmission of altruism is $\varphi$, then the condition for evolution of altruism in the case of sib-to-sib altruism is \citep[Eq.~16]{feldman1985gene}
\begin{equation} \label{eq:feldman1985}
c < b \cdot \varphi - \frac{1-\varphi}{\varphi} \,.
\end{equation}
In this case $\varphi$ takes the role of the relatedness $r$ in \autoref{eq:hamilton_rule}, but the effective benefit $b\cdot \varphi$ is further reduced by $(1-\varphi)/\varphi$.

Non-vertical cultural transmission may be either viewed as horizontal or oblique: horizontal transmission occurs between individuals from the same generation, while oblique transmission occurs  to  offspring from the generation to which their parents belong. 
Evolution under either of these transmission models  can be be more rapid than under pure vertical transmission~\citep{cavalli1981cultural,lycett2008questions,ram2018evolution}.
\citet{lewin2017microbes} and \citet{lewin2020rockpaperscissors} have demonstrated that non-vertical transmission, mediated by microbes that manipulate their host's behavior, can help to explain the evolution of cooperative behavior. Some of their analysis can be applied to cultural transmission, because models of cultural transmission are mathematically similar to those for transmission of infectious diseases~\citep{cavalli1981cultural}.
Importantly, their results depended on non-vertical transmission of the cooperation-determining microbes.

We therefore hypothesize that non-vertical cultural transmission can enhance the evolution of cooperation.
To test this hypothesis we suggest a model in which behavioral changes are mediated by cultural transmission that can occur during social interactions. That is, if there exists assortative meeting in the choice of the social interaction partner, or assortative learning in choice of the cultural transmission partner.
For example, if an individual interacts with a cooperative individual, it might learn that cooperation is an advantageous behavior and will be cooperative in the future. 
Our cultural evolution models include both vertical and non-vertical transmission of cooperation, and we investigate these models using mathematical analysis and simulations.  
Our results demonstrate that cultural transmission can facilitate the evolution of cooperation even when genetic transmission can not, and that treatment of cooperation as a cultural, rather than a genetic, trait can lead to a better understanding of its evolutionary dynamics.

@@

Horizontal cultural transmission occurs between pairs of individuals from the same generation. 
It occurs between social partners with probability $\alpha$, or between a random pair with probability $1-\alpha$ (see~\autoref{fig:horizontal}).
The assortment parameter $\alpha$ is therefore the fraction of population that receives (horizontal transmission) from the social interaction partner, and $1-\alpha$ receives randomly.
Horizontal transmission is not always successful, as one partner may reject the other's phenotype. The probability for successful horizontal transmission of phenotypes $A$ and $B$ are $T_A$ and $T_B$, respectively (\autoref{table:interactions}).

@@

%\begin{enumerate} % p* version
%\item Fixation of cooperation,
%	if $T_A>T_B$ and $0<\tilde{p}^*$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0>\tilde{p}^*$, or
%	if $T_A<T_B$ and $1<\tilde{p}^*$.
%\item Global fixation of defection,
%	if $T_A>T_B$ and $\tilde{p}^*>1$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0<\tilde{p}^*$, or
%	if $T_A<T_B$ and $\tilde{p}^*<0$
%\item Co-existence of both phenotypes at $\tilde{p}^*$, 
%	if $T_A<T_B$ and $0<\tilde{p}^*<1$.
%\end{enumerate}
\begin{enumerate} % gamma version
\item \emph{Fixation of cooperation}:
	if \emph{(i)} $T=T_A=T_B$ and $c < b \cdot \frac{\alpha T}{1-T}$; or
	if \emph{(ii)} $T_A>T_B$ and $c<\gamma_1$; or
	if \emph{(iii)} $T_A<T_B$ and $c<\gamma_2$.
\item \emph{Fixation of defection}:
	if \emph{(iv)} $T=T_A=T_B$ and $c > b \cdot \frac{\alpha T}{1-T}$; or
	if \emph{(vi)} $T_A>T_B$ and $\gamma_2<c$; or
	if \emph{(vi)} $T_A<T_B$ and $\gamma_1<c$.
\item \emph{Fixation of either phenotype depending on initial frequency}: 
	if \emph{(vii)} $T_A>T_B$ and $\gamma_1<c<\gamma_2$.
\item \emph{Coexistence of both phenotypes at $\tilde{p}^*$}:
	if \emph{(viii)} $T_A<T_B$ and $\gamma_2<c<\gamma_1$.
\end{enumerate}
\end{result}

@@

In general, these conditions cannot be formulated in the form of Hamilton's rule ($c<b\cdot r$) due to the horizontal transmission bias $T_A-T_B$.
Without horizontal transmission bias, i.e., with $T_A=T_B$, these conditions reduce to the following  form of Hamilton's rule.\\

\begin{corollary}[Symmetric horizontal transmission]
If $T=T_A=T_B$, then cooperation will take over the population if
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \frac{\alpha T}{1-T}.
\end{equation}
\end{corollary}
Inequality \ref{eq:equal_transmission} is obtained by setting $T_A=T_B$ in inequality \ref{eq:vert_hori_global_condition} and can be interpreted as a version of Hamilton's rule (inequality \ref{eq:hamilton_rule}), where $\alpha T/(1-T)$ can be regarded as the `effective relatedness'.
\autoref{fig:results}a demonstrates this condition. 
\\

\begin{corollary}[No assortment of transmission and cooperation]
If $\alpha=0$ and there is horizontal bias for cooperation ($T_A>T_B$) and
(1) the cost is low compared to the bias ($c<(T_A-T_B)/(1-T_B)$), then cooperation will fix from any positive frequency; or
(2) the cost is low compared to the benefit ($c<(1+b)(T_A-T_B)(1-T_B)$), then cooperation will fix if the initial frequency is high enough ($\tilde{p}_0 > \tilde{p}^*$).
\end{corollary}

@@

\begin{corollary}[Perfect assortment of transmission and cooperation]
When $\alpha=1$, there are only two equilibria, $\tilde{p}=0$ and $\tilde{p}=1$.
The condition for evolution of cooperation (i.e. global stability of $\tilde{p}=1$) is found from inequality \ref{eq:vert_hori_alpha1_condition_proof}, namely
\begin{equation}\label{eq:vert_hori_alpha1}
c < \frac{b \cdot T_A + (T_A - T_B)}{1-T_B}.
\end{equation}
\end{corollary}
With perfect assortment, in inequality \ref{eq:vert_hori_alpha1_condition_proof}
 horizontal transmission  occurs together with the cooperative interaction. The same occurs in \citet{lewin2017microbes}, and therefore this corollary is equivalent to their result (see their eq.~1).

In terms of the cost boundaries, inequality~\ref{eq:vert_hori_alpha1} is equivalent to $c<\gamma_1$, and if $T_A>T_B$ then that suffices for fixation of cooperation. If $T_B>T_A$ then $\gamma_2(\alpha=1)<0$ and again, inequality~\ref{eq:vert_hori_alpha1} is sufficient for increase in the frequency of $A$.
Inequality~\ref{eq:vert_hori_alpha1} can be written as
\begin{equation} \label{eq:vert_hori_alpha1_effective}
1 - (1-c)(1-T_B) < (1+b) T_A ,
\end{equation}
which provides an interesting interpretation for the success of cooperation. 
In the interaction between a cooperator and a defector $(1-c)(1-T_B)$ is the probability that the cooperator remains cooperative and also reproduces. 
Therefore, $1 - (1-c)(1-T_B)$ is the probability that either the cooperator becomes a defector, or that it fails to reproduce.
This is the effective cost for cooperation from this interaction, while $(1+b) T_A$ is the probability that the defector becomes cooperative and reproduces, which is the effective benefit for cooperation from this interaction.
Thus inequality~\ref{eq:vert_hori_alpha1} entails that cooperation can evolve if the effective cost for cooperation is less than the effective benefit.

@@

\begin{corollary}[Intermediate assortment of transmission and cooperation]
The cooperative phenotype will increase from rarity if the assortment is high enough, or specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}

\autoref{fig:results2}c demonstrates these conditions. With horizontal bias for cooperation ($T_A>T_B$; positive x-axis) cooperation fixes in the population from any initial positive frequency (green) if $\alpha$ (y-axis) is above the orange line ($a_1$), or if initially present at a high enough frequency (blue) if $\alpha$ is above the blue line ($a_2$).
With horizontal bias for defection ($T_B>T_A$; negative x-axis) cooperation fixes if $alpha$ is above the blue line ($a_2$), but can be maintained in coexistence with defection (yellow) if $\alpha$ is above the orange line ($a_1$). Without horizontal bias ($T_A=T_B$) fixation occurs if $\frac{c}{b} \cdot \frac{1-T}{T} < \alpha$ (inequality~\ref{eq:equal_transmission}).

@@

In this case $0<v<1$, and the recursion system is more complex,
and we focus on local rather than on global stability.
To proceed, we note that 
Eq.\ \ref{eq:horizontal} can give $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$.
%Equation 4 does not give $\hat{p}$---MF
Eq.\ \ref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\tilde{p}$, since $\hat{p}$ is given in Eq.\ \ref{eq:horizontal} as a function of $\tilde{p}$ and 
Eq.\ \ref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$ (shown in Appendix~\autoref{sec:appendixA}).
We then determine the equilibria, namely, solutions of $\hat{p}' = \hat{p}$, and analyse their local stability.
%: an equilibrium $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})=\bar{w}(\hat{p}'-\hat{p})$ at the equilibrium is negative, $f'(\hat{p}^*)<0$.

@@

We redefine the cost boundaries,
the \emph{cost boundaries},
\begin{equation} \label{eq:cost_boundaries_v}
\begin{aligned}
\hat\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)}, \quad
\hat\gamma_2 = \frac{b v \alpha T_B + (1+b) (T_A - T_B)}{v(1-T_B) + (1-v)(T_A-T_B)},
\end{aligned}
\end{equation}
and define a vertical transmission threshold,
\begin{equation} \label{eq:v_threshold}
\hat v = \frac{T_B - T_A}{1-T_A} \,.
\end{equation}

@@

\begin{result}[Vertical, oblique, and horizontal transmission of cooperation] \label{result:vert_obli_hori}
With vertical, horizontal, and oblique transmission, the cultural evolution of a cooperator phenotype will follow one of the following scenarios depending on the cost boundaries $\hat\gamma_1$ and $\hat\gamma_3$ (\autoref{eq:cost_boundaries_v}) and the vertical threshold $\hat v$ (\autoref{eq:v_threshold}) :

% in beta
%\begin{enumerate}
%\item \emph{Fixation of cooperation}: 
%	if \emph{(i)} $T=T_A=T_B$ and $c < b\cdot \frac{\alpha T}{1-T}$; or
%	if \emph{(ii)} $T_A>T_B$ and $0<\beta_3$; or 
%	if \emph{(iii)} $T_A<T_B$ and $\beta_1<\beta_3$.
%\item \emph{Fixation of the defection}: 
%	if \emph{(iv)}  $T=T_A=T_B$ and $c > b\cdot \frac{\alpha T}{1-T}$; or 
%	if \emph{(v)} $T_A>T_B$ and $\beta_1<\beta_3<0$; or 
%	if \emph{(vi)} $T_A<T_B$ and $\beta_3<0$.
%\item \emph{Coexistence of both phenotypes at $\hat{p}^*$}: 
%	if \emph{(vii)} $T_A < T_B$ and $0<\beta_3<\beta_1$.
%\item \emph{Fixation of either phenotype depending on initial frequency}:
%	if \emph{(viii)}  $T_A>T_B$ and $\beta_3<\beta_1<0$.
%\end{enumerate}

% in gamma
\begin{enumerate}
\item \emph{Fixation of cooperation}: 
	if \emph{(i)} $T=T_A=T_B$ and $c < b\cdot \frac{\alpha T}{1-T}$; or
	if \emph{(ii)} $T_A>T_B$ and $c < \hat\gamma_1$; or 
	if \emph{(iiia)} $T_A<T_B$ and $v>\hat v$ and $c < \hat\gamma_2$; or
	if \emph{(iiib)} $T_A<T_B$ and $v<\hat v$ and $c > \hat\gamma_2$.
\item \emph{Fixation of the defection}: 
	if \emph{(iv)}  $T=T_A=T_B$ and $c > b\cdot \frac{\alpha T}{1-T}$; or 
  if \emph{(v)} $T_A>T_B$ and $\hat{\gamma_1} < c < \hat{\gamma_2}$; or 
	if \emph{(vi)} $T_A<T_B$ and $c>\hat{\gamma_1}$.
\item \emph{Coexistence of both phenotypes at $\hat{p}^*$}: 
  if \emph{(viia)} $T_A < T_B$ and $v<\hat{v}$ and $c<\hat{\gamma_1}$ and $c<\hat{\gamma_2}$; or
  if \emph{(viib)} $T_A < T_B$ and $v>\hat{v}$ and $c<\hat{\gamma_1}$ and $c>\hat{\gamma_2}$. %$0<\beta_3<\beta_1$.
\item \emph{Fixation of either phenotype depending on initial frequency}:
  if \emph{(viii)}  $T_A>T_B$ and $c > \hat{\gamma_2}$ and $c > \hat{\gamma_1}$ and $v>\hat{v}$.
\end{enumerate}
\end{result}

>>>>

% Title page
\title{Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation}

@@

% Abstract
\begin{abstract}
We construct models for the cultural evolution of cooperation in a well-mixed population with vertical, horizontal, and oblique transmission. Conditions are found for fixation and coexistence of cooperation and defection. Comparisons are drawn with Hamilton's rule and the concepts of relatedness and assortment.
%A spatial model is constructed and its results are compared to the well-mixed model.
We show that assortment between cooperation and horizontal transmission facilitates the evolution of cooperation.
% and increases the population mean fitness. 
% TODO what about oblique?
\end{abstract}

@@

\citet{Eshel1982} studied a related model for the evolution of cooperative behavior.
Their model included \emph{assortative meeting}, or non-random encounters, where a fraction $m$ of individuals in the population each interact with an individual of the same phenotype, and a fraction $1-m$ interacts  with a randomly chosen individual.  
Such assortative meeting may be due, for example, to population structure or active partner choice.
In their model, cooperative behavior can evolve if\footnote{In an extended model, which allows an individual to encounter $N$ individuals before choosing a partner, the righthand side is multiplied by $E[N]$, the expected number of encounters \citep[eq.~4.6]{Eshel1982}.
} 
\citep[eq.~3.2]{Eshel1982}
\begin{equation} \label{eq:eshel1982}
c < b \cdot m \,,
\end{equation}
where $b$ and $c$ are the benefit and cost of cooperation. 
Here $m$ in inequality~\ref{eq:eshel1982} takes the role of the relatedness $r$ in inequality~\ref{eq:hamilton_rule}.

The role of assortment in the evolution of altruism was studied by \citet{Fletcher2009assortment} in a \emph{public goods} game.
They found that for altruism to evolve, cooperative individuals must experience more cooperation, on average, than defecting individuals, and ``thus, the evolution of altruism requires (positive) assortment between focal \emph{cooperative} players and cooperative acts in their interaction environment.''
This condition can be summarized by \citep[eq.~2.3]{Fletcher2009assortment}
\begin{equation} \label{eq:fletcher2009}
c < b \cdot (e_C - e_D ) \,,
\end{equation}
where $e_C$ is the probability that a cooperator receives help, and $e_D$ is the probability that a defector receives help.\footnote{inequality~\ref{eq:fletcher2009} generalizes inequality~\ref{eq:hamilton_rule} and inequality~\ref{eq:eshel1982} by substituting $e_C=r + p$, $e_D=p$ and $e_C=m + (1-m)p$, $e_D=(1-m)p$, respectively, where $p$ is the frequency of cooperators.}

Here we study the evolution of a cooperative behavior that is subject to \emph{cultural transmission}, which allows an individual to acquire attitudes or behavioral traits from other individuals in its social group through imitation, learning, or other modes of communication \citep{cavalli1981cultural,richerson2008not}.
\citet{feldman1985gene} introduced the first model for the evolution of altruism by cultural transmission.
They demonstrated that if the fidelity of cultural transmission of altruism is $\varphi$, then the condition for evolution of altruism in the case of sib-to-sib altruism is \citep[Eq.~16]{feldman1985gene}
\begin{equation} \label{eq:feldman1985}
c < b \cdot \varphi - \frac{1-\varphi}{\varphi} \,.
\end{equation}
Here $\varphi$ takes the role of `relatedness' ($r$ in Eq.~\ref{eq:hamilton_rule}) or `assortment' ($m$ in Eq.~\ref{eq:eshel1982}), but the effective benefit $b\cdot \varphi$ is further reduced by $(1-\varphi)/\varphi$. This shows that under a combination of genetic and vertical (parent-to-offspring) cultural transmission, the condition for the evolutionary success of altruism entails a modification of Hamilton's rule (\autoref{eq:hamilton_rule}).

Cultural transmission may be either viewed as vertical, horizontal or oblique:  vertical transmission occurs between parents and offspring, horizontal transmission occurs between individuals from the same generation, and oblique transmission occurs  to offspring from the generation to which their parents belong (i.e. from non-parental adults). 
Evolution under either of these transmission models can be be more rapid than under pure vertical transmission~\citep{cavalli1981cultural,lycett2008questions,ram2018evolution}.
Both \citet{woodcock2006significance} and \citet{lewin2017microbes} demonstrated that non-vertical transmission can help explain the evolution of cooperative behavior (the former using simulations with cultural transmission, the latter using a model where cooperation is mediated by microbes that manipulate their host's behavior.) 
Specifically, some of the analyses by \citet{lewin2017microbes} can be applied to cultural transmission, because models of cultural transmission are mathematically similar to those for transmission of infectious diseases~\citep{cavalli1981cultural}.

Here we hypothesize that non-vertical cultural transmission can enhance the evolution of cooperation because it generates assortment between cooperative individuals.
To test this hypothesis we construct a model in which behavioral changes can be mediated by cultural transmission that can occur during social interactions.
For example, an individual that interacts with a cooperative individual might learn that cooperation is a valuable behavior, and will therefore teach its offspring to be cooperative. 
Our cultural evolution models include both vertical and non-vertical transmission of cooperation, and we investigate these models using mathematical analysis and simulations.  
Our results demonstrate that cultural transmission can facilitate the evolution of cooperation even when genetic transmission can not because it facilitates the generation of assortment, and that treatment of cooperation as a cultural trait, rather than a genetic one, can lead to a better understanding of its evolutionary dynamics.

@@

Horizontal cultural transmission occurs between pairs of individuals from the same generation. 
It occurs between social partners with probability $\alpha$, or between a random pair with probability $1-\alpha$ (see~\autoref{fig:horizontal}).
The assortment rate $\alpha$ is therefore the fraction of population that receives (horizontal transmission) from the social interaction partner, and $1-\alpha$ receives randomly.
Horizontal transmission is not always successful, as one partner may reject the other's phenotype. The probability for successful horizontal transmission of phenotypes $A$ and $B$ are $T_A$ and $T_B$, respectively (\autoref{table:interactions}).

@@

%\begin{enumerate} % p* version
%\item Fixation of cooperation,
%	if $T_A>T_B$ and $0<\tilde{p}^*$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0>\tilde{p}^*$, or
%	if $T_A<T_B$ and $1<\tilde{p}^*$.
%\item Global fixation of defection,
%	if $T_A>T_B$ and $\tilde{p}^*>1$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0<\tilde{p}^*$, or
%	if $T_A<T_B$ and $\tilde{p}^*<0$
%\item Co-existence of both phenotypes at $\tilde{p}^*$, 
%	if $T_A<T_B$ and $0<\tilde{p}^*<1$.
%\end{enumerate}
\begin{enumerate} % gamma version
\item \emph{Fixation of cooperation}:
	if \emph{(i)} $T=T_A=T_B$ and $c < b \cdot \alpha \frac{T}{1-T}$; or
	if \emph{(ii)} $T_A>T_B$ and $c<\gamma_1$; or
	if \emph{(iii)} $T_A<T_B$ and $c<\gamma_2$.
\item \emph{Fixation of defection}:
	if \emph{(iv)} $T=T_A=T_B$ and $c > b \cdot \alpha \frac{T}{1-T}$; or
	if \emph{(vi)} $T_A>T_B$ and $\gamma_2<c$; or
	if \emph{(vi)} $T_A<T_B$ and $\gamma_1<c$.
\item \emph{Fixation of either phenotype depending on initial frequency}: 
	if \emph{(vii)} $T_A>T_B$ and $\gamma_1<c<\gamma_2$.
\item \emph{Coexistence of both phenotypes at $\tilde{p}^*$}:
	if \emph{(viii)} $T_A<T_B$ and $\gamma_2<c<\gamma_1$.
\end{enumerate}
\end{result}

@@

In general, these conditions cannot be formulated in the form of Hamilton's rule ($c<b\cdot r$) due to the horizontal transmission bias $T_A-T_B$.
Without horizontal transmission bias, i.e., with $T=T_A=T_B$, these conditions reduce to a form of Hamilton's rule.\\

\begin{corollary}[Symmetric horizontal transmission]
If $T=T_A=T_B$, then cooperation will take over the population if
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \alpha \frac{T}{1-T}.
\end{equation}
\end{corollary}
Inequality \ref{eq:equal_transmission} is obtained by setting $T=T_A=T_B$ in inequality \ref{eq:vert_hori_global_condition} and can be interpreted as a version of Hamilton's rule (inequality~\ref{eq:hamilton_rule}) or as a version of inequality~\ref{eq:fletcher2009}, where $\alpha T/(1-T)$ can be regarded as the `effective relatedness' or `effective assortment', respectively.
\autoref{fig:results}a demonstrates this condition. 
\\

\begin{corollary}[No assortment of transmission and cooperation]
Without assortment ($\alpha=0$) and with horizontal bias for cooperation ($T_A>T_B$), then 
(1) if the cost is low compared to the bias ($c<(T_A-T_B)/(1-T_B)$), then cooperation will fix from any positive frequency; 
(2) if the cost is low compared to the benefit ($c<(1+b)(T_A-T_B)(1-T_B)$), then cooperation will fix if the initial frequency is high enough ($\tilde{p}_0 > \tilde{p}^*$).
\end{corollary}

@@

\begin{corollary}[Perfect assortment of transmission and cooperation]
With perfect assortment ($\alpha=1$), there are only two equilibria, $\tilde{p}=0$ and $\tilde{p}=1$.
The condition for evolution of cooperation (i.e. global stability of $\tilde{p}=1$) is found from inequality \ref{eq:vert_hori_alpha1_condition_proof}, namely
\begin{equation}\label{eq:vert_hori_alpha1}
c < \frac{b \cdot T_A + (T_A - T_B)}{1-T_B}.
\end{equation}
\end{corollary}
With perfect assortment horizontal transmission always co-occurs with the cooperative interaction. The same occurs in \citet{lewin2017microbes}, and therefore this corollary is equivalent to their result (see their eq.~1).

In terms of the cost boundaries, inequality~\ref{eq:vert_hori_alpha1} is equivalent to $c<\gamma_1$, and if $T_A>T_B$ then that suffices for fixation of cooperation. If $T_B>T_A$ then $\gamma_2(\alpha=1)<0$ and again, inequality~\ref{eq:vert_hori_alpha1} is sufficient for increase in the frequency of $A$.
Inequality~\ref{eq:vert_hori_alpha1} can be written as
\begin{equation} \label{eq:vert_hori_alpha1_effective}
1 - (1-c)(1-T_B) < (1+b) T_A ,
\end{equation}
which provides an interesting interpretation for the success of cooperation. 
In the interaction between a cooperator and a defector, $(1-c)(1-T_B)$ is the probability that the cooperator remains cooperative and also reproduces. 
Therefore, $1 - (1-c)(1-T_B)$ is the probability that either the cooperator becomes a defector, or that it fails to reproduce.
This is the \emph{effective cost of cooperation} from this interaction, while $(1+b) T_A$ is the probability that the defector becomes cooperative and reproduces, which is the \emph{effective benefit of cooperation} from this interaction.
Thus, inequality~\ref{eq:vert_hori_alpha1} entails that cooperation can evolve if the effective cost of cooperation is less than the effective benefit.

@@

\begin{corollary}[Intermediate assortment of transmission and cooperation]
The cooperative phenotype will increase from rarity if the assortment rate is high enough, or specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}

\autoref{fig:results2}c demonstrates these conditions. With horizontal bias for cooperation ($T_A>T_B$; positive x-axis) cooperation fixes in the population from any initial positive frequency (green) if $\alpha$ (y-axis) is above the orange line ($a_1$), or if initially present at a high enough frequency (blue) if $\alpha$ is above the blue line ($a_2$).
With horizontal bias for defection ($T_B>T_A$; negative x-axis) cooperation fixes if $\alpha$ is above the blue line ($a_2$), but can be maintained in coexistence with defection (yellow) if $\alpha$ is above the orange line ($a_1$). Without horizontal bias ($T_A=T_B$) fixation occurs if the assortment rate is high enough, $\alpha > \frac{c}{b} \cdot \frac{1-T}{T}$ (inequality~\ref{eq:equal_transmission}).

@@

With both vertical and oblique transmission, $0<v<1$, the recursion system is more complex,
and we focus on local rather than on global stability.
To proceed, we note that 
Eq.\ \ref{eq:horizontal} can give $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$.
%Equation 4 does not give $\hat{p}$---MF
Eq.\ \ref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\tilde{p}$, since $\hat{p}$ is given in Eq.\ \ref{eq:horizontal} as a function of $\tilde{p}$ and 
Eq.\ \ref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$ (shown in Appendix~\autoref{sec:appendixA}).
We then determine the equilibria, namely, solutions of $\hat{p}' = \hat{p}$, and analyse their local stability.
%: an equilibrium $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})=\bar{w}(\hat{p}'-\hat{p})$ at the equilibrium is negative, $f'(\hat{p}^*)<0$.

@@

% TODO uncomment when Result 3 in terms of c is fixed
%We define new cost boundaries, $\hat\gamma_1$ and $\hat\gamma_2$,
%\begin{equation} \label{eq:cost_boundaries_v}
%\begin{aligned}
%\hat\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)}, \quad
%\hat\gamma_2 = \frac{b v \alpha T_B + (1+b) (T_A - T_B)}{v(1-T_B) + (1-v)(T_A-T_B)},
%\end{aligned}
%\end{equation}
%and define a vertical transmission threshold,
%\begin{equation} \label{eq:v_threshold}
%\hat v = \frac{T_B - T_A}{1-T_A} \,.
%\end{equation}

@@

\begin{result}[Vertical, oblique, and horizontal transmission of cooperation] \label{result:vert_obli_hori}
With vertical, horizontal, and oblique transmission, the cultural evolution of a cooperator phenotype will follow one of the following scenarios depending on the 
polynomial coefficients $\beta_1$ and $\beta_3$ (\autoref{eq:polynomial_coefficients}):
%cost boundaries $\hat\gamma_1$ and $\hat\gamma_3$ (\autoref{eq:cost_boundaries_v}) and the vertical threshold $\hat v$ (\autoref{eq:v_threshold}) :

% in beta
\begin{enumerate}
\item \emph{Fixation of cooperation}: 
	if \emph{(i)} $T=T_A=T_B$ and $c < b\cdot \frac{\alpha T}{1-T}$; or
	if \emph{(ii)} $T_A>T_B$ and $0<\beta_3$; or 
	if \emph{(iii)} $T_A<T_B$ and $\beta_1<\beta_3$.
\item \emph{Fixation of the defection}: 
	if \emph{(iv)}  $T=T_A=T_B$ and $c > b\cdot \frac{\alpha T}{1-T}$; or 
	if \emph{(v)} $T_A>T_B$ and $\beta_1<\beta_3<0$; or 
	if \emph{(vi)} $T_A<T_B$ and $\beta_3<0$.
\item \emph{Coexistence of both phenotypes at $\hat{p}^*$}: 
	if \emph{(vii)} $T_A < T_B$ and $0<\beta_3<\beta_1$.
\item \emph{Fixation of either phenotype depending on initial frequency}:
	if \emph{(viii)}  $T_A>T_B$ and $\beta_3<\beta_1$.
\end{enumerate}

% in gamma - TODO fix
%\begin{enumerate}
%\item \emph{Fixation of cooperation}: 
%	if \emph{(i)} $T=T_A=T_B$ and $c < b\cdot \alpha \frac{T}{1-T}$; or
%	if \emph{(ii)} $T_A>T_B$ and $c < \hat\gamma_1$; or 
%	if \emph{(iii.a)} $T_A<T_B$ and $v>\hat v$ and $c < \hat\gamma_2$; or
%	if \emph{(iii.b)} $T_A<T_B$ and $v<\hat v$ and $\hat\gamma_2 < c$.
%\item \emph{Fixation of the defection}: 
%	if \emph{(iv)}  $T=T_A=T_B$ and $c > b\cdot \alpha \frac{T}{1-T}$; or 
%  if \emph{(v)} $T_A>T_B$ and $\hat\gamma_1 < c < \hat\gamma_2$; or 
%	if \emph{(vi)} $T_A<T_B$ and $\hat\gamma_1 < c$.
%\item \emph{Fixation of either phenotype depending on initial frequency}:
%  if \emph{(viii)} $T_A>T_B$ and $v > \hat{v}$ and $\hat\gamma_2 < c$.
%\item \emph{Coexistence of both phenotypes at $\hat{p}^*$}: 
%  if \emph{(vii.a)} $T_A < T_B$ and $v < \hat{v}$ and $\hat\gamma_1 < c < \hat\gamma_2$; or
%  if \emph{(vii.b)} $T_A < T_B$ and $v>\hat{v}$ and $\hat\gamma_1 < c$ and $\hat\gamma_2 < c$.
%\end{enumerate}

\end{result}

%This result illustrates the crucial role of vertical and oblique transmission: the same cost of cooperation $c$ can lead to remarkably different outcomes if the vertical transmission rate $v$ is below or above the threshold $\hat v$.
% TODO example with figure

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Discussion
\section*{Discussion}
We hypothesized that non-vertical transmission can explain the evolution of cooperation and evaluated this hypothesis using a deterministic discrete-time evolutionary model with fitnesses in the form of payoffs from a prisoner's dilemma game.
Under oblique and horizontal cultural transmission, a horizontal transmission bias for the cooperative phenotype was found to be sufficient and necessary for evolution of cooperation (Result~\autoref{result:obli_hori}).
Under horizontal and vertical cultural transmission, cooperation or defection can either fix or coexist at a stable polymorphism, depending on the relationship between the cost and benefit of cooperation, the horizontal bias, and the assortment between cooperation and transmission (Result~\autoref{result:vert_hori}).
Under a combination of vertical, oblique, and horizontal transmission the dynamics are more complicated, but all outcomes can still occur, depending on parameter values (Result~\autoref{result:vert_obli_hori}).
Remarkably, stable coexistence between cooperation and defection can be maintained if horizontal transmission is biased for defection ($T_B>T_A$) and the cost of cooperation and assortment ratio are intermediate (yellow areas \autoref{fig:results2}).
% TODO figure for result 3
% TODO something about v-hat
